---
title: "p8105_hw2_mkr2158"
author: "Matariya Rattanapan"
UNI: "mkr2158"
output: github_document
date: "2025-09-30"
---

Loading tidyverse

```{r}
library(tidyverse)
library(readxl)
```

## Problem 1

Cleaning up the pols-month dataset.

```{r}
pols_df = 
  read_csv("fivethirtyeight_datasets/pols-month.csv", na = c("NA", ".", "")) |> 
  janitor::clean_names() |> 
  separate(mon, into = c("year", "month", "day"), sep = "-") |> 
  mutate(
    month = recode(month, "01" = "January", "02" = "February", "03" = "March", "04" = "April", "05" = "May", "06" = "June", "07"= "July", "08" = "August", "09" = "September", "10" = "October", "11" = "November", "12" = "December"),
    president = case_when(
      prez_gop == 1 ~ "gop",
      prez_dem == 1 ~ "dem"
    )) |> 
  select(-prez_dem, -prez_gop, -day)
```

Cleaning up the snp dataset.

```{r}
snp_df =
  read_csv("fivethirtyeight_datasets/snp.csv", na = c("NA", ".", "")) |> 
  janitor::clean_names() |> 
  separate(date, into = c("month", "day", "year"), sep = "/") |> 
  mutate(
    month = recode(month,"1" = "January", "2" = "February", "3" = "March", "4" = "April", "5" = "May", "6" = "June", "7"= "July", "8" = "August", "9" = "September", "10" = "October", "11" = "November", "12" = "December"),
    year = case_when(
      as.numeric(year) >= 50 ~ paste0("19", year),
      as.numeric(year) < 50 ~ paste0("20", year)
    )
  ) |> 
  arrange(year, month) |> 
  select(year, month, close)
```

Cleaning up unemployment dataset.

```{r}
unemploy_df =
  read_csv("fivethirtyeight_datasets/unemployment.csv", na = c("NA", ".", "")) |> 
  janitor::clean_names() |> 
  pivot_longer(
    cols = jan:dec,
    names_to = "month",
    values_to = "unemployment_pct"
  ) |> 
  mutate(
    month = recode(month, "jan" = "January", "feb" = "February", "mar" = "March", "apr" = "April", "may" = "May", "jun" = "June", "jul" = "July", "aug" = "August", "sep" = "September", "oct" = "October", "nov" = "November", "dec" = "December"),
    year = as.character(year)
  ) |> 
  arrange(year, month) |> 
  select(year, month, unemployment_pct)
```

Merging the cleaned up pols-month, snp, and unemployment datasets. 

```{r}
merged_df =
  left_join(pols_df, snp_df, by = c("year", "month"))

finalmerge_df =
  left_join(merged_df, unemploy_df, by = c("year", "month"))
```

Using the `knitr::kable` function to check to see our final merged data was created successfully.

```{r}
knitr::kable(head(finalmerge_df, 5))
```

The dataset from `pols_df` was cleaned up and contains `r nrow(pols_df)` observations and `r ncol(pols_df)` variables. It contains important variables describing the number of democratic or republican politicians at the associated years and months, including the presidents, governors, senators, and representatives. The dataset from `snp_df` was cleaned up and contains `r nrow(snp_df)` observations and `r ncol(snp_df)` variables. It contains the closing value of the Standard & Poor's (S&P) stock market index at the associated years and months. The dataset from `unemployment` was cleaned up and contained `r nrow(unemploy_df)` observations and `r ncol(unemploy_df)` variables. It contains the percentage of unemployment at the associated years and months, respectively. Our final dataset `finalmerge_df` contains `r nrow(finalmerge_df)` observations and `r ncol(finalmerge_df)` variables. The range of year for our final merged dataset is from January 1947 to June 2015. Key variables in our final dataset include the `year`, `month`, `president`, `close`, and `unemployment_pct`.



## Problem 2

Read and clean the Mr. Trash Wheel sheet

```{r}
mrtrash_df =
  read_excel("Trash_Wheel_Collection_Data.xlsx", sheet = "Mr. Trash Wheel", range = "A2:N709",na = c("NA", ".", ""), skip = 1) |> 
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |> 
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    wheel = "mr"
  ) |> 
  rename(weight = weight_tons, volume = volume_cubic_yards, cigarette = cigarette_butts)
```

Read and clean Professor Trash Wheel sheet

```{r}
proftrash_df =
  read_excel("Trash_Wheel_Collection_Data.xlsx", sheet = "Professor Trash Wheel", range = "A2:M709",na = c("NA", ".", ""), skip = 1) |> 
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |> 
  mutate(
    wheel = "prof"
  ) |> 
  rename(weight = weight_tons, volume = volume_cubic_yards, cigarette = cigarette_butts)
```

Read and clean Gwynnda Trash Wheel sheet

```{r}
gwynnda_df =
  read_excel("Trash_Wheel_Collection_Data.xlsx", sheet = "Gwynns Falls Trash Wheel", range = "A2:L709",na = c("NA", ".", ""), skip = 1) |> 
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |> 
  mutate(
    wheel = "gwynns"
  ) |> 
  rename(weight = weight_tons, volume = volume_cubic_yards, cigarette = cigarette_butts)
```




